Propositions as Sessions, Mechanically

### 2015-04-07 paper/background.tex
Added beginnings of CP propositions and processes.
Need to ask Simon what he thinks about all these figures.
Some further discussion of permutation issues.
Perhaps an example of permutation?

\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. GV presents session types using a modified version
of the standard definition presented in \S~\ref{sec:pre}. Figure~\ref{??}
presents the types of GV with session types ranged over by $S$. Note that
\lstinline{end} has been replaced with $\outend$ and $\inend$ denoting the
terminating session type for $\cdots$ and $\cdots$ respectively \todo{what are
  the interpretations for $\outend$ and $\inend$? Need a figure for GV
  sessions}. CP presents linear logical connectives with session type
interpretations inspired by
\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in Figure~\ref{??}
\todo{figure for CP types with session type
  interpretation}. \citeauthor{Wadler:2014} describes a continuation-passing
style (CPS) translation from GV session types to CP classical linear logic
propositions.  \todo{terms and interpretations for CP/GV?} The process
calculus CP is slightly different to $\pi$DILL by
\citeauthor{Caires:2010:STI}, apart from being classical in nature, the
constructs for server replication are modified; CP provides separate weakening
and contraction rules whereas these are encoded in one rule for process
reduction in $\pi$DILL. Additionally, rather than a two-sided sequent
calculus, \citeauthor{Wadler:2014} uses one-sided sequents for defining CP,
which provides a more intuitive presentation of session type duality and a
clearer correspondence to a $\lambda$-calculi surface syntax, simplifying the
translation presentation. As presented, GV does not support all features of CP
and thus the translation is one way (GV to
CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
provide support for polymorphism and replication, following a similar CPS
translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

>>>>

\input{cp-types-terms}

\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
process calculus, CP, based on the earlier work by
\citeauthor{Caires:2010:STI}, and a functional language, GV, based on LAST by
\citeauthor{Gay:2010:LAST}. Figure~\ref{fig:gvtnts} presents the types and
terms GV with types ranged over by $T$, session types ranged over by $S$ and
terms ranged over by $L$, $M$ and $N$. GV session types are a modified version
of the definition presented in \S~\ref{sec:pre} where \lstinline{end} has been
replaced with $\outend$ and $\inend$. The intention is that $\outend$
($\inend$) terminates a session type whose last operation is an output
(input), and they correspond to the units of the respective linear logical
connectives \todo{Do I need to elaborate on the translation? Yet another
  figure?}. CP provides linear logical connectives with session type
interpretations inspired by
\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in
Figure~\ref{fig:cptnts} along with the terms of CP. The process calculus CP is
slightly different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from
being classical in nature, the constructs for server replication are modified:
CP provides separate weakening and contraction rules whereas these are encoded
in one rule for process reduction in $\pi$DILL. Additionally, rather than a
two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
for defining CP, which provides a more intuitive presentation of session type
duality and a clearer correspondence to a $\lambda$-calculi surface syntax
like GV. \citeauthor{Wadler:2014} describes a continuation-passing style (CPS)
translation from GV types and terms to CP classical linear logic propositions
and processes, respectively. As presented, GV does not support all features of
CP and thus the translation is one way (GV to
CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
provide support for polymorphism and replication following a similar CPS
translation scheme. Lastly,
\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
semantics for GV and their system has greater similarities with LAST than
prior GV works, for example, by treating send and receive as linear functions
rather than language primitives.

### 2015-04-07 paper/formalisation.tex
Moved adoption and focus to future work section. Tweaking background.
Some further discussion of permutation issues.
Perhaps an example of permutation?

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a simply typed lambda calculus this is not onerous,
but GV has four binding constructs, CP has six and propositions have
two. Indeed, a lot of effort was expended setting up this
infrastructure. However, I believe one of the contributions of this work is to
articulate what one needs from state of the art metatheory libraries in order
to handle session-based languages, and more generally, linear type systems.

@@

propositions

processes

@@

Formalising the process calculus CP provided many challenges for the
metatheory library since to my knowledge, the library has never been applied
to a process calculus before. So for instance, in the lambda calculus one
typically does not work up to permutation of binders, whereas in a process
calculus these kinds of structural equivalences are standard. Additionally,
substitution and opening are only defined for names. In other words, one
cannot substitute an arbitrary process term for a name, unlike in the lambda
calculus. Thus frameworks which assume a lambda calculi style operations are
not immediately amenable to the CP setting~\cite{Lee:2012}. Ordering in
environments is ignored so the development required lemmas allowing the
permutation of environments. Support for permutating environments is not
currently supported by the Metatheory library but some general lemmas were
added as a result of this work. Additionally, custom tactics were developed to
automatatically solve some simple permutation goals, however these tactics are
not a ``catch all'' and some generalising still needs to be done.

@@

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form (ignoring
permutation premises). In other words bound variables are represented using de
Brujin indices and free variables are represented using an abstract type
provided by the Metatheory library. The rules with binders reflect this by
specifying the type of a bound variable in the constructor rather than the
name. Cofinite quantification is performed in the hypotheses to these rules by
excluding a finite set of names from opening the subprocesses. Note that the
rules for polymorphism, $\exists$ and $\forall$, were included in the final
Coq development since GV (\S~\ref{sec:gv}) does not support polymorphism
making the rules unnecessary for the translation. They appear in the Figure
simply for the sake of completeness. Additionally, as noted by
\citeauthor{Lee:2012}, support for polymorphism in type variables creates an
explosion of the infrastructure requiring duplication of opening and
substituting definitions and lemmas.

@@

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients. contraction is not supported because
of the difficulty in specifying its principal cut elimination rule. In order
to specify the rule one requires to show the following:

\todo{v A -> !0(A).P | Q{x/x'}
  ==>
v A -> !0(A).P | v A -> !0(A).P | contract (Gamma) Q}

Unfortunately, the reduction relation does not have knowledge of the
environment typing process P. This fact does not impact upon expressing
weakening since for weakening only the names (not the bindings) within the
environment are required. The paper presentation does not specify where
weakening could occur; it is tacitly assumed weakening can always be pushed
further up the derivation tree e.g. AxCut rule. a consequence of this choice
is that for the empty choice rule there may only be a single binding in the
context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}).

@@

The principal cut elimination rules are presented in Figure~\ref{??}. These
rules are expressed to indicate some additional complexity not present in the
original presentation. For example, structural equivalences and associativity
are not unique and different process terms result depending on initial binder
permutation. ${0 <~> 1}P$ indicates the permuting of binders $0$ and $1$ in
process P. The paper presentation specifies only one possible associativity
rule lending to the specific split of the linear environments. The reduction
rules were complicated to prove due to being composed of multiple constructor
instances and differed significantly to the standard operational semantics one
typically deals with in lambda calculus formalisms.
linear environments caused a problem with the splitting up of variables.

@@

% ... subject reduction....

@@

types

terms
\end{comment}

>>>>

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a simply typed $\lambda$-calculus ($\stlc$) this is
not onerous, but GV has four binding constructs, CP has six and propositions
have two. Indeed, a lot of effort was expended setting up this
infrastructure. However, I believe one of the contributions of this work is to
articulate what one needs from state of the art metatheory libraries in order
to handle session-based languages, and more generally, linear type systems.

@@

@@

@@

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
tedious details relating to environment uniqueness and ordering (see below)
elided. In other words, bound variables are represented using de Brujin
indices and free variables are represented using an abstract type provided by
the Metatheory library. The rules with binders reflect this by specifying the
type of a bound variable in the constructor rather than the name. Cofinite
quantification is performed in the hypotheses to these rules by excluding a
finite set of names from opening the subprocesses where opening a process $P$
with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
$\exists$ and $\forall$, were included in the final Coq development since GV
(\S~\ref{sec:gv}) does not support polymorphism making the rules unnecessary
for the translation. Further, as noted by \citeauthor{Lee:2012}, support for
polymorphism in type variables creates an explosion of the infrastructure
requiring additional of opening and substituting definitions and lemmas. They
appear in the Figure simply for the sake of completeness.

Formalising the process calculus CP provided many challenges for the
Metatheory library since to my knowledge, the library has never been applied
to a process calculus before, and the treatment of names, binding and
substitution differ to that present in $\lambda$-calculi. For instance, in the
$\stlc$ one typically does not work up to permutation of binders, whereas in a
process calculus these kinds of structural equivalences are
standard. Additionally, substitution and opening are only defined for
names. In other words, one cannot substitute an arbitrary process term for a
name, unlike in the $\lambda$-calculus. Thus frameworks which assume
$\lambda$-calculi style operations are not immediately amenable to the CP
setting~\cite{Lee:2012}. Ordering in environments is ignored so the
development required lemmas allowing the permutation of
environments. Permutating environments is not currently supported by the
Metatheory library but some general lemmas were added as a result of this
work, and custom tactics were developed to automatically solve certain
permutation goals via simple proof search. However, there are still many cases
where manual sequences of primitive lemmas, such as appending, commutativity
and transitivity, need to be used. A lot of these cases are similar but there
would not appear to be an advantage in developing a tactic to do the
work. Doing so would create a dependency between a hard-to-debug and obscure
Ltac definition and the proofs which utilise it, becoming a burden to
change. This development philosophy treats Ltac definitions akin to macros, a
shorthand for expressing an oft-repeated sequence of tactics, and it is
doubtful much can be gained from it unless the use of Ltac definitions can be
made more transparent.

@@

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients. contraction is not supported because
of the difficulty in specifying its principal cut elimination rule. In order
to specify the rule one requires to show the following: \todo{fix binders of
  Q}
\begin{gather*}
\tm{\nu A.(!0(A).P \mid ?z[z',z''].Q)}
\quad \becomes \quad \\
\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
\end{gather*}

where $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
$\tm{Q}$ with the names in $\Gamma$.

Unfortunately, the reduction relation does not have knowledge of the
environment typing process P. This fact does not impact upon expressing
weakening since for weakening only the names (not the bindings) within the
environment are required. The paper presentation does not specify where
weakening could occur; it is tacitly assumed weakening can always be pushed
further up the derivation tree e.g. AxCut rule. a consequence of this choice
is that for the empty choice rule there may only be a single binding in the
context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}). one can compare alms with this version of cp
since it weakening rule corresponds to affine logic

@@

The structural equvialences and principal cut elimination rules are presented
in Figures~\ref{??} and \ref{??} respectively. These rules are expressed to
indicate some additional complexity not present in the original
presentation. For example, structural equivalences are not unique and
different process terms result depending on initial binder permutation. ${0
  <~> 1}P$ indicates the permuting of binders $0$ and $1$ in process P. The
paper presentation specifies only one possible associativity rule lending to
the specific split of the linear environments. The reduction rules were
complicated to prove due to being composed of multiple constructor instances
and differed significantly to the standard operational semantics one typically
deals with in lambda calculus formalisms.  linear environments caused a
problem with the splitting up of variables.

@@

% ... subject reduction ...

@@

### 2015-04-07 paper/introduction.tex
Some further discussion of permutation issues.
Perhaps an example of permutation?

This project describes work towards the formalisation of calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types (GV)
using interactive theorem proving (in particular, the system Coq). The
formalisation is a contribution to the practice of programming language
metatheory and informs what is required of a framework for formalising lambda
calculus type systems with session types. Such a framework should enable a
whole class of related languages and type systems to be formalised with little
extra proof effort and is inspired by previous work on a mechanised framework
for $\pi$-calculus type systems~\cite{Gay:2001:FFP}. The current work focuses
on the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, taking advantage of previous
proof effort through the re-use of existing libraries for programming language
metatheory. The work has presented a number of interesting challenges for the
formalisation of programming languages and as a result generated a number of
useful guidelines and issues to consider.

>>>>

This project describes work towards the formalisation of calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types (GV)
using interactive theorem proving (in particular, the system Coq). The
formalisation is a contribution to the practice of programming language
metatheory and informs what is required of a framework for formalising
$\lambda$-calculus type systems with session types. Such a framework should
enable a whole class of related languages and type systems to be formalised
with little extra proof effort and is inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}. The
current work focuses on the $\lambda$-calculus to provide a basis on which
session types for mainstream programming languages can be studied, taking
advantage of previous proof effort through the re-use of existing libraries
for programming language metatheory. The work has presented a number of
interesting challenges for the formalisation of programming languages and as a
result generated a number of useful guidelines and issues to consider.
